\chapter*{Abstract}
\addcontentsline{toc}{chapter}{Abstract}

The exponential growth of the Hilbert space with the number of constituents renders many quantum-information tasks intractable for exact classical methods, motivating the search for scalable approximations. This thesis develops two physics-informed deep learning frameworks that address such tasks by embedding the known physical structure of each problem directly into the network architecture and the training objective, so that the learned solutions remain consistent with the underlying quantum mechanics.

The first framework concerns the dynamics of closed quantum systems governed by time-dependent Hamiltonians, whose time-ordered evolution operator admits no closed-form solution. Given the time-evolution operator on a sparse set of sampled times, a Physics-Informed Neural Network interpolates it over the entire time domain without ever being supplied the Hamiltonian explicitly, incorporating unitarity as a physical constraint and, for the larger systems, predicting an effective Hamiltonian propagated through structure-preserving integrators. Across systems of $2$ to $8$ qubits the method attains mean gate fidelities close to $0.99$, reproduces the dynamics accurately even under coarse temporal sampling, and achieves inference speedups of roughly one order of magnitude over a geodesic-interpolation baseline.

The second framework addresses the Quantum Marginal Problem, a QMA-complete task in which one seeks a global density matrix compatible with a prescribed set of quantum marginals. Combining a convolutional denoising autoencoder with the Marginal Imposition Operator, and encoding density matrices as two-channel images with a loss that enforces Hermiticity, positivity, and normalization, the model reconstructs physically valid global states for systems of $3$ to $8$ qubits. A hybrid scheme recovers states whose marginals match the targets with perfect fidelity in many cases, transfer learning extends the model across system sizes with limited retraining, and, once trained, inference is substantially faster than state-of-the-art semidefinite programming solvers, even succeeding in regimes where the latter fail to run.

Together, these results illustrate a common pattern: when the governing physics of a problem is built into the learning process, deep learning becomes not a replacement for physical understanding but an extension of it, capable of addressing problems whose exact treatment lies beyond the reach of conventional computation.
