\chapter*{Acknowledgements}
\addcontentsline{toc}{chapter}{Acknowledgements}

This thesis is the result of a journey that I did not walk alone, and these lines are a small attempt to thank those who walked it with me.

To my wife, Carolina, for her love, her patience, and her unwavering support through every long day and every late night. You have been my calm and my strength throughout this entire process, and none of this would have been the same without you by my side.

To my friends, for the laughter, the encouragement, and the welcome distractions that kept me grounded; and to my mother-in-law, for her generosity and her constant support, which I deeply appreciate.

To my pets, my faithful companions, who kept me company through every working day at home, curling up beside me during the long hours and reminding me, in their own way, to take a break now and then.

To my advisor and to the colleagues who served on my committee, for their guidance, their careful reading, and the many discussions that shaped and sharpened this work. I am grateful for their time, their insight, and their trust.

To the Universidad de Concepción, for supporting me in presenting this work at various conferences and events, an experience that enriched both this thesis and my growth as a researcher; and to ANID, for the doctoral fellowship that made this research possible.

Finally, and most especially, to my mother, may she rest in peace. For her love, her sacrifices, and everything she gave me, this achievement is as much hers as it is mine. Without her I would never have reached where I am today. This thesis is dedicated to her memory.
